Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-31 paper.tex

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0, the current state-of-the-art
coalescent simulator both with respect to efficiency as well as feature-richness.
Here we provide an overview of the extensive abilities of \msprime\ 1.0, 
demonstrate through numerous performance comparisons its advantages over
current, alternative software packages, and finally lay out future plans for
further open-source development of what we hope will become an industry
standard piece of computational infrastructure for the genetics community.

>>>>

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0, a state-of-the-art
coalescent simulator both with respect to efficiency as well as feature-richness.
Here we provide an overview of the extensive abilities of \msprime\ 1.0, 
demonstrate through numerous performance comparisons its advantages over
current, alternative software packages, and finally lay out future plans for
further open-source development of what we hope will become an industry
standard piece of computational infrastructure for the genetics community.
